---
title: "Gov 1372 - Hierarchies"
author: "Paddy Adams"
date: "September 30, 2021"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) # filter(), group_by(), summarize
library(readr) # read_csv()
library(ggplot2) # ggplot()
library(estimatr) # difference_in_means()
library(skimr)
library(gt)
library(broom)
library(ggcorrplot)

```

# Social Dominance and Politics

Welcome to the fourth Data Exploration Assignment. This week, we read about Social Dominance Theory, and its related psychological construct, Social Dominance Orientation (SDO). SDO is measured through a survey scale consisting of 16 items that we already explored in class. In this assignment, you will explore data on SDO and its relationship with other variables from a nationally representative survey fielded in 2018.

Note that the actionable part of each question is **bolded**. 

**Data Details:**

* File Name: \texttt{sdo\_data.csv}

* Source: These data are condensed and adapted from a [survey](https://wthh.dataforprogress.org/get-the-data) by Data for Progress in 2018 (N = 3144). The data are representative of 2018 voters.

Variable Name         | Variable Description
--------------------- | --------------------------------------
\texttt{sdo5}       | Five-point social dominance orientation (SDO) scale: 1. Minimum SDO; ...; 5. Maximum SDO
\texttt{female}         | Indicator for whether or not the respondent is female. Coded 1 if respondent is female, 0 otherwise.
\texttt{birthyr}         | Respondent's birth year
\texttt{educ}        | Education: 1. Didn't graduate HS; 2. HS graduate; 3. Some college; 4. 2-year college; 5. 4-year college; 6. Postgraduate degree
\texttt{race}  | Race: 1. White; 2. Black or African-American; 3. Hispanic or Latino; 4. Asian or Asian-American; 5. Native American; 6. Mixed Race; 7. Other; 8. Middle Eastern
\texttt{favor\_trump}    | Favorability of Donald Trump: 1. very unfavorable; ...; 4. very favorable
\texttt{favor\_blm}    | Favorability of Black Lives Matter: 1. very unfavorable; ...; 4. very favorable
\texttt{favor\_metoo}        | Favorability of the Me Too movement: 1. very unfavorable; ...; 4. very favorable
\texttt{american\_customs}       | "The growing number of newcomers from other countries threatens traditional American customs and values": 1. Strongly disagree; ...; 5. Strongly agree
\texttt{race\_ident}       | "How important is being [respondent's race] to you?": 1. Not at all important; ...; 4. Very important
\texttt{pid3}       | Three-category party identification: 1. Democrat; 2. Independent; 3. Republican
\texttt{ideo5}       | Five-category political ideology: 1. Very liberal; ...; 5. Very conservative
\texttt{fear\_of\_demographic\_change}       | Fear of demographic change in the US: 0. Least fearful; ...; 1. Most fearful
\texttt{confederate\_flag}       | Is the Confederate flag mostly a symbol of slavery and white supremacy or Southern heritage and culture? Coded either "slavery" or "heritage"
\texttt{presvote16}       | Vote choice in the 2016 presidential election

## Question 1: REQUIRED

Before looking at data, the science of political psychology often involves building surveys. The teaching team builds the surveys you take using an online survey-building software called Qualtrics. This is often the same software that researchers use to build surveys and collect data that is eventually published in peer-reviewed journals. In this question, you'll create your own brief survey. 

**THIS QUESTION IS REQUIRED FOR ALL STUDENTS. Go to harvard.qualtrics.com and log in using your HarvardKey. Click "Create new project", then select "Survey". You can name your survey whatever you like. Leave the other two drop-down options at their default and click "Create project". Now you can input the SDO scale, which is given below. Make sure to include all 16 items, split into two sub-scales, in your survey. They are split into two sub-scales here, but they don't need to be in your survey. For each item, there should be seven response categories: Strongly favor, Somewhat favor, Slightly favor, Neutral, Slightly oppose, Somewhat oppose, Strongly oppose. Think about the format you think is best for these questions, available under "Question Type". How might the format of the questions affect the responses you get from the survey, or the experience respondents have while taking the survey?  Also consider question ordering and how that may also affect the responses. BE SURE TO UPLOAD A SCREENSHOT OF YOUR QUALTRICS SURVEY TO YOUR BLOG THIS WEEK.**

**Dominance Sub-Scale**
\begin{enumerate}
  \item Some groups of people must be kept in their place.
  \item It's probably a good thing that certain groups are at the top and other groups are at the bottom.
  \item An ideal society requires some groups to be on top and others to be on the bottom.
  \item Some groups of people are simply inferior to other groups.
  \item Groups at the bottom are just as deserving as groups at the top. 
  \item No one group should dominate in society.
  \item Groups at the bottom should not have to stay in their place.
  \item Group dominance is a poor principle.
\end{enumerate}

**Anti-Egalitarianism Sub-Scale**
\begin{enumerate}
  \item We should not push for group equality.
  \item We shouldn't try to guarantee that every group has the same quality of life.
  \item It is unjust to try to make groups equal.
  \item Group equality should not be our primary goal.
  \item We should work to give all groups an equal chance to succeed.
  \item We should do what we can to equalize conditions for different groups.
  \item No matter how much effort it takes, we ought to strive to ensure that all groups have the same chance in life.
  \item Group equality should be our ideal.
\end{enumerate}

## Question 2

Now let's take a look at the data.
```{r, message=FALSE}

sdo <- read_csv("sdo_data.csv")

```

**What is the distribution of social dominance orientation in the sample? Make a plot, and report the mean and standard deviation of SDO in the sample. Extend this problem by splitting the plot by party ID of the respondent. Comment on what you find.**


```{r}

hist_plot <- sdo %>% 
  ggplot(aes(x = sdo5)) +
  geom_histogram(aes(y = ..density..), bins = 10,  fill = "darkgreen") +
  stat_function(fun = dnorm, args = list(mean = mean(sdo$sdo5), sd = sd(sdo$sdo5))) +
  theme_classic() 

hist_plot

```


Social Dominance Orientation, or 'SDO', is a personality metric that gives a rough idea to what extent they support hierarchy and dominance over lower-status social groups. The dataset for this weeks blog gives just over 3000 survey responses that allow us to see how this measure of SDO relates to other socio-political metrics such as political ideology, age, sex and favourability of current political issues. From this we can explore what groups have higher social status. 

To begin with, it's interesting to explore the most broad implications of SDO and how it relates to partisanship. As SDO is a predisposition toward anti-egalitarianism within and between groups, one would expect that republicans would have a higher SDO score. In this survey partisanship is given by the 'pid3' metric, which has 3 options - '1' designating a Democrat, '2' being an Independent and '3' being for a Republican. When plotted against SDO we can see the differences:


```{r}

sdo_avg <- sdo %>% 
  filter(is.na(pid3) == FALSE) %>% 
  group_by(pid3) %>% 
  summarise(avg = mean(sdo5), sd = sd(sdo5))

sdo_avg %>% 
  mutate(pid3 = c("Democrat", "Independent", "Republican")) %>% 
  gt() %>% 
  tab_header(title = "SDO Mean and Standard Deviation") %>% 
  cols_label(pid3 = "Party",
             avg = "Mean",
             sd = "SD") %>%
  fmt_number(columns = 2:3, decimals = 2)


```


```{r}

pid3_names <- list(
  '1' = "Democrat",
  '2' = "Independent",
  '3' = "Republican")

labeller <- function(variable,value){
  return(pid3_names[value])
}

sdo$pid3 <- as.factor(sdo$pid3)

pid3_hist_plot <- sdo %>% 
  filter(is.na(pid3) == FALSE) %>% 
  ggplot(aes(x = sdo5)) +
  geom_histogram(aes(fill = pid3), bins = 10, alpha = 0.5) +
  theme_classic() +
  facet_grid( ~ pid3, labeller = labeller) +
  geom_vline(data = sdo_avg, aes(xintercept = avg), color = 'darkgreen') +
  scale_fill_manual(values = c("blue", "grey", "red")) +
  labs(title = "Social Dominance Orientation by Party Identification",
       x = "SDO",
       y = "Count",
       subtitle = "Mean Shown by Verticle Line") +
  theme(legend.position = "none") 
  

pid3_hist_plot


```


There is quite a clear distinction here between the three political associations, with the simplest way to see this being to look at where the means of the three groups lie. It is interesting that the SDO count '1' is quite so high amongst the Democrat respondents, with neither of the other groups having such a dominant mode. I find this intriguing as I would have thought spread would be more consistent with the other two and not have quite such a dominant SDO score. One possible explanation would be to consider that a lot of questions on the survey have a 'moral highground' that many Democrat respondents may have considered and have had weighing on their conscience when responding instead of giving genuine answers - similar to the heuristics and biases we looked at a couple of weeks ago. 





## Question 3

In the reading for this week, we saw that gender is central to social dominance theory, which predicts that men tend to have higher SDO than women do. **Is this true in this sample as well? Report the average SDO for men and women. Comment on what you find. Extend this question by reporting the difference in means along with the p-value. Is the difference significant at a .05 significance level?**

```{r}

v <- sdo %>% 
  group_by(female) %>% 
  summarise(mean = mean(sdo5)) %>% 
  mutate(female = c("Male", "Female"))

chisq.test(v$female, v$mean, correct=FALSE)



```



To further explore this data I wanted to look at the correlations between SDO and favourability of a couple of political issues also in this dataset:  


## Question 4
**What is the correlation between `sdo5` and the `favor_trump` variable? Is the correlation statistically different from zero? You can use cor.test() for this question. Interpret what you find. If you want, extend this question by creating a scatterplot with the line of best fit to visualize the relationship.  You can use geom_point() in the ggplot architecture for this.**

```{r}

cor.test(sdo$sdo5, sdo$favor_trump)
cor.test(sdo$sdo5, sdo$favor_blm)

sdo$favor_blm <- as.numeric(sdo$favor_blm)

sdo %>% 
  filter(is.na(pid3) == FALSE) %>% 
  ggplot(aes(x = favor_trump, y = sdo5, color = favor_blm)) +
  geom_jitter() +
  geom_smooth(method = 'lm', se = FALSE, color = 'hotpink') +
  theme_classic() +
  labs(title = "SDO and Trump Support",
       x = "Trump Favourability",
       y = "SDO",
       color = "BLM Favourability\n") 



```






The correlation between SDO and whether the respondent favoured Trump is 0.5946, showing quite a strong positive correlation that is in line with what we would expect considering how SDO related to partisanship and Trump's anti-egalitarian policies. Similarly the correlation between support for the BLM movement (-0.5979) shows that as SDO increases support for the movement goes down; again in line with the theory of SDO. I'd be interested in whether this is less correlation and more causation however, as the rhetoric around the BLM movement was mostly negative from Trump and thus if one were to support and favour Trump they would be less likely to support BLM. Another interesting thing is to see which respondents have given outlying results, such as those who are strongly against Trump and pro-BLM but nevertheless voted Republican. Alternatively those who voted Democrat and yet favoured Trump and where anti-BLM. These may be anomalies, but would be interesting to look at further nonetheless. 




## Question 5

Correlation matrices, like the one below, are useful for visualizing the pairwise relationships between several variables. They allow you to see the correlation coefficients of of many relationships at once. **Plot a correlation matrix of the correlation between SDO and some of the variables you think might be related to SDO and to each other. Choose at least 3 variables in addition to SDO. Before you make your plot, briefly discuss why you think the variables might all be related. The package `ggcorrplot` may be useful here. Discuss what you see in your plot.**

```{r}

c <- sdo %>% 
  select(c(1:6, 12)) %>% 
  cor()

ggcorrplot(c)

```



