---
title: "Gov 1372 - Hierarchies"
author: "Paddy Adams"
date: "September 30, 2021"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) # filter(), group_by(), summarize
library(readr) # read_csv()
library(ggplot2) # ggplot()
library(estimatr) # difference_in_means()
library(skimr)
library(gt)
library(broom)
library(ggcorrplot)

```

# Social Dominance and Politics

```{r, message=FALSE, include=FALSE}

sdo <- read_csv("sdo_data.csv")

```

Social Dominance Orientation, or 'SDO', is a personality metric that gives a rough idea to what extent they support hierarchy and dominance over lower-status social groups. The dataset for this weeks blog gives just over 3000 survey responses that allow us to see how this measure of SDO relates to other socio-political metrics such as political ideology, age, sex and favourability of current political issues. From this we can explore what groups have higher social status. 

To begin with, it's interesting to explore the most broad implications of SDO and how it relates to partisanship. As SDO is a predisposition toward anti-egalitarianism within and between groups, one would expect that republicans would have a higher SDO score. In this survey partisanship is given by the 'pid3' metric, which has 3 options - '1' designating a Democrat, '2' being an Independent and '3' being for a Republican. When plotted against SDO we can see the differences:


```{r,echo=F, warning=FALSE, message=FALSE}

sdo_avg <- sdo %>% 
  filter(is.na(pid3) == FALSE) %>% 
  group_by(pid3) %>% 
  summarise(avg = mean(sdo5), sd = sd(sdo5))

sdo_avg %>% 
  mutate(pid3 = c("Democrat", "Independent", "Republican")) %>% 
  gt() %>% 
  tab_header(title = "SDO Mean and Standard Deviation") %>% 
  cols_label(pid3 = "Party",
             avg = "Mean",
             sd = "SD") %>%
  fmt_number(columns = 2:3, decimals = 2)


```


```{r, echo=FALSE, warning=FALSE}

pid3_names <- list(
  '1' = "Democrat",
  '2' = "Independent",
  '3' = "Republican")

labeller <- function(variable,value){
  return(pid3_names[value])
}

sdo$pid3 <- as.factor(sdo$pid3)

pid3_hist_plot <- sdo %>% 
  filter(is.na(pid3) == FALSE) %>% 
  ggplot(aes(x = sdo5)) +
  geom_histogram(aes(fill = pid3), bins = 10, alpha = 0.5) +
  theme_classic() +
  facet_grid( ~ pid3, labeller = labeller) +
  geom_vline(data = sdo_avg, aes(xintercept = avg), color = 'darkgreen') +
  scale_fill_manual(values = c("blue", "grey", "red")) +
  labs(title = "Social Dominance Orientation by Party Identification",
       x = "SDO",
       y = "Count",
       subtitle = "Mean Shown by Verticle Line") +
  theme(legend.position = "none") 
  

pid3_hist_plot


```


There is quite a clear distinction here between the three political associations, with the simplest way to see this being to look at where the means of the three groups lie. It is interesting that the SDO count '1' is quite so high amongst the Democrat respondents, with neither of the other groups having such a dominant mode. I find this intriguing as I would have thought spread would be more consistent with the other two and not have quite such a dominant SDO score. One possible explanation would be to consider that a lot of questions on the survey have a 'moral highground' that many Democrat respondents may have considered and have had weighing on their conscience when responding instead of giving genuine answers - similar to the heuristics and biases we looked at a couple of weeks ago. 


To further explore this data I wanted to look at the correlations between SDO and favourability of a couple of political issues also in this dataset:  


```{r, include=FALSE, echo=FALSE}
cor.test(sdo$sdo5, sdo$favor_trump)
cor.test(sdo$sdo5, sdo$favor_blm)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}


sdo$favor_blm <- as.numeric(sdo$favor_blm)

sdo %>% 
  filter(is.na(pid3) == FALSE) %>% 
  ggplot(aes(x = favor_trump, y = sdo5, color = favor_blm)) +
  geom_jitter() +
  geom_smooth(method = 'lm', se = FALSE, color = 'hotpink') +
  theme_classic() +
  labs(title = "SDO and Trump Support",
       x = "Trump Favourability",
       y = "SDO",
       color = "BLM Favourability\n") 



```


The correlation between SDO and whether the respondent favoured Trump is 0.5946, showing quite a strong positive correlation that is in line with what we would expect considering how SDO related to partisanship and Trump's anti-egalitarian policies. Similarly the correlation between support for the BLM movement (-0.5979) shows that as SDO increases support for the movement goes down; again in line with the theory of SDO. I'd be interested in whether this is less correlation and more causation however, as the rhetoric around the BLM movement was mostly negative from Trump and thus if one were to support and favour Trump they would be less likely to support BLM. Another interesting thing is to see which respondents have given outlying results, such as those who are strongly against Trump and pro-BLM but nevertheless voted Republican. Alternatively those who voted Democrat and yet favoured Trump and where anti-BLM. These may be anomalies, but would be interesting to look at further nonetheless. 




