---
title: "Gov 1372 - Groups and Identities"
author: "Paddy Adams"
date: "September 23, 2021"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(skimr)


```



```{r, include=FALSE, message=F}

data <- read_csv('Sep23ClassData_clean.csv')

```



This weeks class survey and resulting data looks at the partisan divide and affective polarization of the group. Following the same method to that taken by Klar et al. (2018) we can establish the individual affective polarization by looking at how the respondent feels about their child marrying either in-party or out-party. This, combined with how partisan the individual is, gives an idea as to their polarization. 


```{r, echo=FALSE}


ggplot(data = data %>% filter(is.na(outPartyUnhappy)==FALSE)) + 
  geom_bar(mapping = aes(x = polarized, y = ..prop.., group = 1), stat = "count", fill = "darkgreen") + 
  facet_wrap(~treatment) + 
  ylab("Proportion") + 
  xlab("Affectively Polarized") +
  theme_classic() 

```


By looking at just the 'true' columns we can compare this data to that collected by Klar et al. with relative ease. While the broad pattern is the same with 'rarely' showing the lowest total and 'frequently' the highest, the overall proportion of those showing polarization is much higher. There is also a much bigger range in the proportion that are affectivly polarized when we compare between those who talk about politics 'rarely' and those who do 'frequently', but I would suggest this has something to do with the demographic of those surveyed as the 'rarely' result compares very similarly to the proportion found in the original paper. To further explain how the demographic plays an important role here it is important to note that this is a class of people who understandably both talk a lot and posses strong opinions about politics (this is a Government class after all). These strong opinions amongst most of the group explains why the control and frequently group are much more affectively polarized. My assumption here being that most of the class hold strongly democrat/liberal views (72 out if the 82 put their party as Democrat) and would be fairly unhappy with their child marrying out-party. Compared to the general population, who probably care less about politics even if they talk about it more than 'rarely', this presents as a much lower level of affective polarization.  

Another noticeable question is as to why those who talk about politics the most seem to be less polarized than those in the control group. I think the fact that 'rarely' ranks lower than than the others is fairly self-explanatory, as often discussing politics is a taboo subject so it doesn't matter so much what your potential daughter/son-in-law decides with their politics. With regard to why those who talk about politics the most aren't the most polarized, I would presume that this is because those who are the most engaged will know exactly where they disagree or agree with the other party and are thus more open to debate and confident in conversation with out-party relations. They are able to enjoy healthy partisan debate. Alternatively they conform to the idea of “negative partisan-ship” (Abramowitz and Webster 2016) whereby they actively dislike many in-party, but support their party out of a stronger dislike of the other party (for example those who would have voted for Bernie Sanders would say they didn't vote for Biden rather they voted against Trump). As such they feel less strongly about out-party marriage of their children and are less affectivley polarized. 





